The force of mortality by life lived is the force of increment by life left in stationary populations

### 2015-03-05 LifetableIncrement/ReplyReviewer2.tex

\documentclass{article}

\usepackage{amsmath}
\usepackage{natbib}
\bibpunct{(}{)}{,}{a}{}{;} 
\newcommand{\dd}{\; \mathrm{d}}
\usepackage{url}

\begin{document}
\title{Second reply to reviewers of submission 2518: ``The force of mortality by
life lived is the force of increment by life left in stationary populations''}
\author{Author Redacted}
\maketitle

My thanks to the reviewers for the corrections and constructive suggestions.
\textbf{Reviewer A} found an error that owes to my own misunderstanding of
integrals, which until now I've naively understood as perfectly analogous to discrete sums.
In other words, the old equation (5) works when you draw it on paper, but now I
understand why it does not work with calculus notation. Reviewer A also
provided a solution that saves the argument, and for which I am grateful.

@@

Here's a link to the 1986 paper: 
\url{http://www.persee.fr/web/revues/home/prescript/article/espos_0755-7809_1986_num_4_2_1120}

The key sentence in the  1986 paper, which led me to discover the other more
obscure source, is at the bottom of page 159, and continues into page 161 (pg
160 is a figure page)- he states Careys equality in plain French and cites a 1985 manuscript ``Mouvements et mod\`{e}les de population''
to back it up. I've never been able to locate the 1985 version, but it appears
to be something he maintained over a series of years, and the
first linked manuscript above appears to be from 1989 (so I cite it as such). Looking at the inside title page it says 1986. It appears to have been placed online in 2002.

To settle the date more exactly, one would have to ask N. Brouard when he
thought of it (or where he saw it). I imagine he went to Cameroon in 1985 with a
notebook of concepts to demonstrate, and either he knew it beforehand, or it occurred to him there, but it wasn't, as far as I
can tell, in a distributed print form until 1989. He apparently cited his 1985
workshop materials, though I could be mistaken. I've not seen a hard copy.

Regarding \citet{rao2014generalization}, the citation seems OK to me, but it was
easy to miss since the bibliography is alphabetical by last name but printed
starting with first name (Arni). I'll reformat as the journal requires.

\textbf{Finally}, I added a reference to \citet{chiang1984life}, who provides
an alternative but equivalent expression of $\sigma^2(y|a)$ (in the History \&
Related Results section), and I also fixed some minor typos here and there.

Thanks to all for the careful reviewing

### 2015-04-02 LifetableIncrement/LifetableIncrement.tex
moved sentence, changed integration limits to make less redundant

%%This is a very basic article template.
%%There is just one section and two subsections.
\documentclass{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% header, footers for review
\usepackage{fancyhdr}
\pagestyle{fancy}

%\chead{submitted as a relationship report to be included in the Special
%Collection ``Formal Relationships'' edited by Joshua R. Goldstein and James W.
%Vaupel}
%\lfoot{Submission \#2518 for review}
%\rfoot{}
\fancyhf{}
\fancyhead[C]{\footnotesize{submitted as a relationship report to be included in the
Special Collection ``Formal Relationships'' edited by Joshua R. Goldstein and James W.
Vaupel}}
\fancyfoot[R]{\footnotesize{\bfseries{page \thepage}}}
\fancyfoot[L]{\footnotesize\bfseries{Submission \#2518, revised}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
%\setlength\headheight{117.89105pt}
\setcounter{page}{2}
\pagestyle{fancy}

\makeatletter
\let\ps@plain\ps@fancy 
\makeatother

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsmath}
\usepackage{caption}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{setspace}
\doublespacing
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
% for the d in integrals
\newcommand{\dd}{\; \mathrm{d}}
% end preamble.

@@

\title{The force of mortality by life lived is the force of increment by life
left in stationary populations}
\author{Author Redacted}
\maketitle
\section*{Structured abstract}

@@

The rate of birth accumulation into death cohorts over thanatological age,
$\mu^\star (y)$, is equal to the rate of attrition of birth cohorts
over chronological age, $\mu (a)$ in stationary populations. The stock of deaths in
the future for chronological age $a$, $l(a)$, is symmetrically a description of births in the
past when structured by thanatological age, $l^\star(y)$. The remaining
lifetable columns are subject to similar reinterpretations when viewed under
thanatological age. The main contribution of this relationship is to point out
the symmetry of chronological decrement and thanatological renewal processes in
the case of stationary populations.

@@

The term ``thanatological age'' has not previously appeared in the
literature, and was coined by Ken Wachter sometime prior to 2001. Thanatos was
the Greek god of death, which is used as the reference point for age under this
perspective. Explicit decompositions of chronological
age groups into remaining lifespan classes is to our knowledge only found in
\citet{brouard1986structure}, who redistributed
population pyramids by remaining years of life, and
\citet{miller2001increasing}, who examined medicare expenditure as a function of time until death using the
same lifetable identities. Equivalence of chronological and
thanatological age structures in stationarity, also known as Carey's equality,
is proven by \citet{brouard1989mouvements}, again in
 \citet{vaupel2009life},\footnote{\citet{vaupel2009life} added detail to the
 more summary result reported in \citet{goldstein2009life}, which showed that
 the mean (or total) remaining lifespan is equal to mean (or total) life lived in the stationary population.} and in more detail by \citet{rao2014generalization}.
These latter two papers and \citet{goldstein2012historical} describe more of the
lineage of the result presented here.

The estimation of remaining lifespan has motivated demography since
the invention of the lifetable, but this notion is typically dealt with as a
mean (expectancy). Age-specific remaining lifespan distributions are not often
used to explicitly decompose demographic quantities, such as population counts. Using
the above identities, one may estimate the total population with $y$
remaining years of life, $P(y)$ directly
\begin{equation}
\label{eq:getpy}
P(y) = \int _0 ^\infty P(a)\mu(a+y)\frac{l(a+y)}{l(a)} \dd a \quad \text{,}
\end{equation}
\noindent as \citet{brouard1986structure} did, and so approximate a population's
thanatological age structure according to some mortality assumptions. This result differs from the more common approach
of summing the population within age classes bounded by some values interpolated
along the remaining life expectancy function (e.g.
\citet{sanderson2005average,sanderson2007new,sanderson2010remeasuring}, who
follow the line of \citet{hersch1944demographie} and \citet{ryder1975notes}).

@@

Similarly, taking $f(y|a)$ as the conditional density of remaining
lifespans, one may calculate the variance of age-specific remaining life
spans (thanatological age), where $e(a)$ is remaining life expectancy at age
$a$:
\begin{align}
e(a) =& \frac{\int _{y=0} ^\infty l(a+y) \dd y}{l(a)}
\intertext{The variance of $y$ given survival to age $a$ is}
\sigma ^2 (y|a) =& \int _{y=0}^\infty \left( e(a) - y\right)^2 f(y|a) \dd y
\quad \text{,}
\end{align}
\noindent and this function will have some non-monotonic pattern over age
(in human populations) that remains to be explored.\footnote{This
definiton works out to be the same as \citet{chiang1984life}, Chapter 10,
Equation 6.10.} In the reliability literature, $\sigma ^2(y|a)$ is called the variance residual life function (VRLF), and its properties have been described for various common distributions
\citep[see for example][]{gupta2006variance}.

@@

>>>>

%%This is a very basic article template.
%%There is just one section and two subsections.
\documentclass[leqno]{article}
% leqno is used for equations on the left.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%% To the editors: this part of the preamble was used for markup
%%%%%%%%%% during the review process, but I comment it out now so you have
%%%%%%%%%% an easier time of editing
% header, footers for review
%\usepackage{fancyhdr}
%\pagestyle{fancy}

%\chead{submitted as a relationship report to be included in the Special
%Collection ``Formal Relationships'' edited by Joshua R. Goldstein and James W.
%Vaupel}
%\lfoot{Submission \#2518 for review}
%\rfoot{}
%\fancyhf{}
%\fancyhead[C]{\footnotesize{submitted as a relationship report to be included
% in the Special Collection ``Formal Relationships'' edited by Joshua R. Goldstein and James W.
%Vaupel}}
%\fancyfoot[R]{\footnotesize{\bfseries{page \thepage}}}
%\fancyfoot[L]{\footnotesize\bfseries{Submission \#2518, revised}}
%\renewcommand{\headrulewidth}{0pt}
%\renewcommand{\footrulewidth}{0pt}
%\setlength\headheight{117.89105pt}
%\setcounter{page}{2}
%\pagestyle{fancy}

%\makeatletter
%\let\ps@plain\ps@fancy 
%\makeatother

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsmath}
\usepackage{caption}
\usepackage{placeins}
\usepackage{graphicx}
\usepackage{natbib} % natbib called for bibtex, used with defaults. 
                    % if formatting is eventually done using package arguments,
                    % it'd be helpful for authors to know the combo of arguments
                    % required to meet reference guidelines. Or maybe a style
                    % could be written?
\usepackage{setspace}
\usepackage{authblk} % for affiliation
\doublespacing
\bibpunct{(}{)}{,}{a}{}{;} 
\usepackage{url}
% for the d in integrals
\newcommand{\dd}{\; \mathrm{d}}
% end preamble.

@@

\title{The force of mortality by life lived is the force of increment by life
left in stationary populations}
\author{Tim Riffe}
\affil{Department of Demography, University of California, Berkeley}
\maketitle
\section*{Structured abstract}

@@

The rate of birth accumulation into death cohorts over thanatological age,
$\mu^\star (y)$, is equal to the rate of attrition of birth cohorts
over chronological age, $\mu (a)$ in stationary populations. The stock of deaths in
the future for chronological age $a$, $l(a)$, is symmetrically a description of births in the
past when structured by thanatological age, $l^\star(y)$. The remaining
lifetable columns are subject to similar reinterpretations when viewed under
thanatological age.

@@

The main contribution of this relationship is to point out
the symmetry of chronological decrement and thanatological renewal processes in
the case of stationary populations. The term ``thanatological age'' has not
previously appeared in the literature, and was coined by Ken Wachter sometime prior to 2001. Thanatos was
the Greek god of death, which is used as the reference point for age under this
perspective. Explicit decompositions of chronological
age groups into remaining lifespan classes is to our knowledge only found in
\citet{brouard1986structure}, who redistributed
population pyramids by remaining years of life, and
\citet{miller2001increasing}, who examined medicare expenditure as a function of time until death using the
same lifetable identities. Equivalence of chronological and
thanatological age structures in stationarity, also known as Carey's equality,
is proven by \citet{brouard1989mouvements}, again in
 \citet{vaupel2009life},\footnote{\citet{vaupel2009life} added detail to the
 more summary result reported in \citet{goldstein2009life}, which showed that
 the mean (or total) remaining lifespan is equal to mean (or total) life lived in the stationary population.} and in more detail by \citet{rao2014generalization}.
These latter two papers and \citet{goldstein2012historical} describe more of the
lineage of the result presented here.

The estimation of remaining lifespan has motivated demography since
the invention of the lifetable, but this notion is typically dealt with as a
mean (expectancy). Age-specific remaining lifespan distributions are not often
used to explicitly decompose demographic quantities, such as population counts. Using
the above identities, one may estimate the total population with $y$
remaining years of life, $P(y)$ directly
\begin{equation}
\label{eq:getpy}
P(y) = \int _0 ^\infty P(a)\mu(a+y)\frac{l(a+y)}{l(a)} \dd a \quad \text{,}
\end{equation}
\noindent as \citet{brouard1986structure} did, and so approximate a population's
thanatological age structure according to some mortality assumptions. This result differs from the more common approach
of summing the population within age classes bounded by some values interpolated
along the remaining life expectancy function (e.g.,
\citet{sanderson2005average,sanderson2007new,sanderson2010remeasuring}, who
follow the line of \citet{hersch1944demographie} and \citet{ryder1975notes}).

@@

Similarly, taking $f(y|a)$ as the conditional density of remaining
lifespans, one may calculate the variance of age-specific remaining life
spans (thanatological age), where $e(a)$ is remaining life expectancy at age
$a$:
\begin{align}
e(a) =& \frac{\int _{0} ^\infty l(a+y) \dd y}{l(a)}
\intertext{The variance of $y$ given survival to age $a$ is}
\sigma ^2 (y|a) =& \int _{0}^\infty \left( e(a) - y\right)^2 f(y|a) \dd y
\quad \text{,}
\end{align}
\noindent and this function will have some non-monotonic pattern over age
(in human populations) that remains to be explored.\footnote{This
definiton works out to be the same as \citet{chiang1984life}, Chapter 10,
Equation 6.10.} In the reliability literature, $\sigma ^2(y|a)$ is called the variance residual life function (VRLF), and its properties have been described for various common distributions
\citep[see for example][]{gupta2006variance}.

@@

\section*{Acknowledgements}
A big thank you to three anonymous reviewers for providing helpful 
suggestions, comments, and corrections. This work was supported by the National Institute On
Aging of the U.S. National Institutes of Health (NIH) under Award Numbers
R01-AG011552 and R01-AG040245. The content is solely the responsibility of the
author and does not necessarily represent the official views of the NIH.
